---
title: "NYC Citi Bike Data Analysis"
author: "Prathiba"
date: "30/08/2021"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# 1. Documentation

#### Domain knowledge (1.1)

Citi Bike is New York City's bike share system, and the largest in the
nation. Citi Bike launched in May 2013 and has become an essential part
of the transportation network. It's fun, efficient and affordable. The
bikes can be unlocked from one station and returned to any other station
in the system, making them ideal for one-way trips.Citi Bike is
available for use 24 hours/day, 7 days/week, 365 days/year. You can read
more about the company the data comes from on the [NYC Citi Bike
website](https://www.citibikenyc.com/about)

#### Business requirements (1.2)

The bike sharing system is becoming more popular in the process of
greener initiative. This report helps to

-   Understand the pattern of bike hires over a period of time to
    identify when it is widely used.

-   Analyse the pattern of bike hires with respect to the rider
    demographics such as age, gender or the type of the user.

-   Identify the geographical spread of the starting points of bike
    hires

#### Business processes and data flow (1.3)

![This diagram was created using draw.io at
<https://app.diagrams.net>](dataflow_nyc_bikes.png)

#### Data visualisation as a tool for decision-making (1.4)

One of the main challenges in the bike sharing system is to effectively
plan the resource usage. This report helps to visualise the traffic
level at each station on a daily basis. Based on the traffic level, the
demands can be predicted and the number of available bikes at that
station can be intepreted or forecasted.

#### Data types (1.5)

| Attributes    | Description                     | Data Types |
|---------------|---------------------------------|------------|
| bike_id       | Unique ID for the bike          | Factor     |
| start_time    | Starting time from the origin   | POSIXct    |
| stop_time     | Ending Time at the destination  | POSIXct    |
| start_station | Origin Station                  | Factor     |
| start_lat     | Origin Station's Latitude       | Number     |
| start_long    | Origin Station's Longitude      | Number     |
| end_station   | Destination Station             | Factor     |
| end_lat       | Destination Station's Latitude  | Number     |
| end_long      | Destination Station's Longitude | Number     |
| type          | Type of user                    | Factor     |
| birth_year    | Birth Year of the user          | Number     |
| gender        | Gender of the user              | Factor     |

#### Data quality and data bias (1.6)

The dataset description says, **" A sample from NYC Citi Bike usage of
10 bikes throughout 2018"** which may result in *selection bias*.
However we assumed that the sample seeks to accurately reflect the
characteristics of the larger group and hence doesn't affect the quality
of the data.

# 2. Data cleaning

#### Preparing data for visualisation (1.7)

```{r message=FALSE, warning=FALSE}
#Load the library
library(tsibbledata)
library(tsibble)
library(feasts)
library(tidyverse)
library(lubridate)
library(leaflet)
library(slider)
```

```{r}
#Load the data
nyc_bikes_df <- nyc_bikes
```

-   The time attributes like month, week, day & year are derived from
    the start time.

```{r}
#Create new columns for Year, Month, Week, Day
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(year = year(start_time),
         month = month(start_time, label = TRUE),
         week = week (start_time),
         day = day(start_time)
         )
```

-   The gender with type "Unknown" are considered as NA as a part of
    cleaning process.

```{r}
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate (gender = na_if(gender, "Unknown")) 
```

```{r}

color_theme <- function() {
  theme(
    plot.background = element_rect(fill ="white"),
    plot.title = element_text(size = rel(1.5), colour = "blue"),
    plot.title.position = "plot",
    
    panel.border = element_rect(colour = "orange", fill = NA, linetype = 1),
    panel.background = element_rect(fill = "white"),
    panel.grid =  element_line(colour = "grey85", linetype = 1, size = 0.5),
    
    axis.text = element_text(colour = "steelblue", face = "italic", size = 10 ),
    axis.title = element_text(colour = "black" , face = "bold", size = 12),
    axis.ticks = element_line(colour = "steelblue"),
    
    legend.box.background = element_rect(),
    legend.box.margin = margin(6, 6, 6, 6)

  )
}
```

# 3. Data visualisation

#### Process and design (2.1, 2.7, 2.8, 2.9)

First, I investigated the bike hire pattern with simple line chart
starting with year and later aggregated with month to identify when it
is widely used. Then I analysed the bike hire pattern with respect to
the rider demographics such as age, gender or the type of the user to
see if there were any relation. I have also represented the geographical
spread using spatial plots with cluster option. I ensured that the key
visualizations are accurate and depicts the clear understanding of the
requirement and not misleading. I also wanted to convey that the plots
were aesthetically pleasing. All visualizations were made in RStudio
using the ggplot and leaflet package.

#### Visualisations (2.2, 2.3, 2.4, 2.5, 2.6)

### [*Understand the pattern of bike hires over a period of time*]{.ul}

Let's first analyse the bike hire pattern over year.

```{r}
#Pattern of bike hire within a year
nyc_bikes_df_rolling <- nyc_bikes_df %>% 
  index_by (date =  as_date(start_time)) %>% 
  summarise(no_of_bike_hire = n()) %>% 
  mutate(
    bike_hire_moving_avg = slide_dbl(
      .x = no_of_bike_hire, 
      .f = ~ mean(., na.rm = TRUE),
      .before = 10,
      .after = 10
    )
  )

ggplot(nyc_bikes_df_rolling) + 
  geom_line(aes(x = date, y = no_of_bike_hire), colour = "grey") + 
  geom_line(aes(x = date, y = bike_hire_moving_avg), colour = "blue")+
  labs( x = "\nYear",
        y = "Number of bike hire",
        title = "Representation of Bike Hire within a Year\n") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")+
  color_theme()

```

The above plot shows the representation of number of bike hire on a
daily basis within a year where the blue line shows the moving average
of the number of bike hire. There seems to be more number of bike hire
between July and October. Let's try to aggregate this to month's data
and visualize further.

```{r}
#Pattern of bike hire over months
nyc_bikes_df_month <- nyc_bikes_df %>% 
  index_by (month) %>% 
  summarise(no_of_bike_hire = n()) %>% 
    ggplot(aes(x = month, y = no_of_bike_hire))+
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "blue") +
  labs( x = "\nMonths",
        y = "Number of bike hire",
        title = "Representation of Bike Hire over months\n")+
  color_theme()
  

nyc_bikes_df_month
```

The line chart above shows the pattern of bike hire over a period of 12
month of the year 2018. As month progresses from left to right, points
connect the number of bike hire . We can read from the general slope of
the line and its vertical positions that the count is improved from
March to August , then dropped gradually to December.

This shows that the bike is hired predominantly from Spring to Autumn,
with highest being the Summer time.

Let's further explore the usage of bike over the weeks. Since the number
of bike hire is predominantly high between June and August, let's
visualise the data for those month,

```{r}
#Pattern of bike hire within a month
nyc_bikes_df_days <- nyc_bikes_df %>% 
  filter (month %in% c("Jun","Jul","Aug")) %>% 
  index_by (date =  as_date(start_time)) %>% 
  mutate(no_of_bike_hire = n()) %>% 
  ggplot() + 
  aes(x = date, y = no_of_bike_hire) +
  geom_line(colour = "steelblue") +
  facet_grid(~month, scales = "free")+
  labs( x = "\nDays",
        y = "Number of bike hire",
        title = "Representation of Bike Hire over Days" )+
  scale_x_date(breaks = "1 week", date_labels = "%d")+
  color_theme()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

nyc_bikes_df_days
```

There is an obvious cyclical pattern, with weekdays (particulary the
beginning of weeks) having about much traffic than weekends. Further
drilling down to hours might give a view on the busiest hours of the day
which might be helpful in further optimization.

### [*Analyse the pattern of bike hires with respect to the rider demographics*]{.ul}

Let's check the hire patterns between the bike riders based on gender
and the type of bike rider.

```{r}
 nyc_bikes_df %>% 
  filter (!(is.na(gender))) %>% 
  index_by(month) %>% 
  group_by(gender) %>% 
  mutate(no_of_bike_hire = n()) %>% 
  ggplot(aes(x = month, y = no_of_bike_hire)) +
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "steelblue") +
  facet_grid(~gender ~ type)+
  labs( x = "\nMonths",
        y = "Number of bike hire",
        title = "Representation of Bike Hire patterns based on Gender and Bike rider")+
  color_theme()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
        
 
```

This line plot is a overall representation of the bike hire patterns
based on gender and type of bike rider over a period of 12 months for
the year 2018.

It gives a brief summary stating

-   Male tend to hire more bikes than female.

-   Subscriber tend to hire more bikes when compared to Customer.

Let's explore which age group hire more bikes

```{r}
nyc_bikes_df %>% 
  mutate (age = year(start_time) - birth_year) %>% 
  index_by(month) %>% 
  summarise(age_mean = mean(age)) %>% 
  ggplot(aes(x = month, y = age_mean)) +
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "steelblue") +
  labs( x = "\nMonths",
        y = "Age",
        title = "Representation of Bike Hire patterns based on demographics",
        subtitle = "(Data taken for the year 2018)")+
  color_theme()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
        
```

The above line chart shows that on an average, riders between age 36 and
38 hire often.

To summarise, though the bike rider's demographics doesn't seems to have
a direct impact on the hire pattern, they are dependent to each other.

### [*Identify the geographical spread of the starting points of bike hires*]{.ul}

Now we understood the maximum usage of the bike hire pattern and
analysed with respect to the rider demographics, let's analyze the
busiest station on a daily basis. It is calculated based on the total
number of bike hire per day. This measure helps to identify the traffic
level of a particular station.

Since there are more bike hires between June and August, Let's visualize
those data using map based display.

```{r}
nyc_bikes_df_filtered <- nyc_bikes_df %>% 
  filter (month %in% c("Jun","Jul","Aug"))

nyc_bikes_df_map <- leaflet(nyc_bikes_df_filtered) %>%
  addTiles() %>%  
  addMarkers(lng = ~start_long, lat = ~start_lat,
             clusterOptions = markerClusterOptions())

nyc_bikes_df_map

```

The map shows the location of each station grouped with markers together
(called as clusters) . Each cluster is identified by its color scale
which really helps to visualize the number of bike hire (check-in) at
that station. The orange circle indicates the busier station with green
being, comparatively lowest traffic level. Based on that, the Jersey
City seems to have the busiest traffic,especially Groove Street (to be
precise). This helps to give an estimate of how to forecasts the future
demand. ie. the minimum available number of bikes at that station.

**Conclusion:**

The bike usage pattern for the year 2018 is analysed over time. Based on
the analysis, the bike hire is most frequently used from Spring to
Autumn, especially in the months between June and August. It is also
observed that the bike usage is more during weekdays than weekends.
Considering the bike rider demographics, male seems to hire more
frequently than the female. Also subscribers hire more frequently when
compared to customers.Finally the busiest station is narrow down using
the clustering analysis from the spatial plot by which the resource
usage can be planned more effectively .

To conclude, these results provide a framework for the business planners
to assess the current process and to develop more prediction analysis on
the number of available bikes at any given station.

**Other Insights:**

-   In our analysis, if we narrow down further to hours the bike was
    hired, it might help to identify the busiest timings of the day. It
    gives a better view to optimize the availability of bikes.

-   Adding weather conditions to our analysis might give a different
    insights on the rider's perspective.
