---
title: "Communicating Data"
author: "Prathiba"
date: "30/08/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The business question says, "What is the pattern of bike hires over time
(e.g. within a year, month, week, or day)?"

Do we need to display the plot of all patterns (like year, month, week
and day) ? Or can we choose the one that is only relevant.

# 1. Documentation

#### Domain knowledge (1.1)

Citi Bike is New York City's bike share system, and the largest in the
nation. Citi Bike launched in May 2013 and has become an essential part
of the transportation network. It's fun, efficient and affordable. The
bikes can be unlocked from one station and returned to any other station
in the system, making them ideal for one-way trips.Citi Bike is
available for use 24 hours/day, 7 days/week, 365 days/year.

#### Business requirements (1.2)

Bike sharing system is becoming more popular in the process of greener
initiative. The highlight of this report is to Understand the pattern of
bike hires over a period of time (from days to year) Analyse whether the
pattern of bike hires changes with respect to the rider demographics
such as age, gender or the type of the trip.

#### Business processes and data flow (1.3)

Create a diagram that shows how the Citi Bike dataset is created (which
variables are added when) and demonstrates an understanding of the
business process and flow of the data.

![This diagram was created using draw.io at
<https://app.diagrams.net>](dataflow_nyc_bikes.png)

#### Data visualisation as a tool for decision-making (1.4)

One of the main challenges in the Bike sharing process is to effectively
plan the resource usage. This report helps to visualise the available
bike hire pattern which in turn allow us to calculate the traffic level
at each station on a daily basis. This can be more useful in the
prediction analysis on the number of available bikes at any given
station.

#### Data types (1.5)

bike_id : Factor start_time : POSIXct stop_time : POSIXct start_station:
Factor start_lat : num start_long : num end_station : Factor end_lat :
num end_long : num type : Factor birth_year : num gender : Factor

#### Data quality and data bias (1.6)

The dataset description says, " A sample from NYC Citi Bike usage of 10
bikes throughout 2018" which may result in selection bias. However we
assumed that the sample seeks to accurately reflect the characteristics
of the larger group and hence doesn't affect the quality of the data.

# 2. Data cleaning

#### Preparing data for visualisation (1.7)

```{r}
#Load the library
library(tsibbledata)
library(tsibble)
library(feasts)
library(tidyverse)
library(lubridate)
```

```{r}
#Load the data
nyc_bikes_df <- nyc_bikes
```

```{r}
#Create new columns for Year, Month, Week, Day
nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(year = year(start_time),
         month = month(start_time, label = TRUE),
         week = week (start_time),
         day = day(start_time)
         )
```

# 3. Data visualisation

#### Process and design (2.1, 2.7, 2.8, 2.9)

Briefly describe your data visualisation process.

For example: *First, I investigated variables of interest with simple
single-variable distribution plots. Then I plotted variables of interest
to the business against each other to see if any patterns were present
in the data. After selecting key visualisations I ensured these were
clearly portraying the message I wanted to convey, were aesthetically
pleasing, as well as accurate and not misleading. All visualisations
were made in RStudio using the `ggplot` package.*

#### Visualisations (2.2, 2.3, 2.4, 2.5, 2.6)

Let's first analyse the bike hire pattern over time.

Pattern of bike hire within a Year

```{r}
nyc_bikes_df_year <- nyc_bikes_df %>% 
  index_by (date =  as_date(start_time)) %>% 
  summarise(count = n()) %>% 
    ggplot() + 
  aes(x = date, y = count) +
  geom_line(colour = "steelblue") +
  labs( x = "\nYear",
        y = "Number of bike hire",
        title = "Representation of Bike Hire within a Year\n")

nyc_bikes_df_year
```

[**Pattern of bike hire within a month**]{.ul}

```{r}
nyc_bikes_df_week <- nyc_bikes_df %>% 
  filter (month %in% c("Jul","Aug")) %>% 
  index_by (week) %>% 
  mutate(count = n()) %>% 
  ggplot() + 
  aes(x = week, y = count) +
  geom_line(colour = "steelblue") +
  facet_grid(~month, scales = "free")+
  labs( x = "\nWeeks",
        y = "Number of bike hire",
        title = "Representation of Bike Hire within a Year\n")

nyc_bikes_df_week

```

```{r}
nyc_bikes_df %>% 
  mutate (age = year(start_time) - birth_year) %>% 
  group_by (gender, type) %>% 
  summarise(age = mean(age)) %>% 
  gg_season(age)+
  xlab("Months")
```

```{r}
library(slider)
nyc_bikes_df_rolling <- nyc_bikes_df %>% 
  mutate (age = year(start_time) - birth_year) %>% 
  mutate(
    bike_hire_moving_avg = slide_dbl(
      .x = age, 
      .f = ~ mean(., na.rm = TRUE),
      .before = 10,
      .after = 10
    )
  )

nyc_bikes_df_rolling


ggplot(nyc_bikes_df_rolling) + 
  geom_line(aes(x = as_date(start_time), y = age), colour = "grey") + 
  geom_line(aes(x = as_date(start_time), y = bike_hire_moving_avg), colour = "red")

```

Let's look at these data individually.

Bike Hire Patterns across gender

```{r}
 nyc_bikes_df %>% 
  group_by (month,gender) %>% 
  mutate(count = n()) %>% 
  index_by (month) %>% 
    ggplot(aes(x = month, y = count))+
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "steelblue") +
  facet_grid(~gender)+
  labs( x = "\nMonths",
        y = "Number of bike hire",
        title = "Representation of Bike Hire across Gender",
        subtitle = "(Over Months)")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

```

```{r}
nyc_bikes_df %>% 
  group_by (month,type) %>% 
  mutate(count = n()) %>% 
  index_by (month) %>% 
    ggplot(aes(x = month, y = count))+
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "steelblue") +
  facet_grid(~type)+
  labs( x = "\nMonths",
        y = "Age (Years)",
        title = "Representation of Bike Hire based on the type of trip",
        subtitle = "(Over Months)")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

```

Bike Hire Patterns based on the age

```{r}
 nyc_bikes_df %>% 
  mutate (age = year(start_time) - birth_year) %>% 
  index_by(month) %>%
  summarise(age_mean = mean(age, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = age_mean)) +
  geom_point(colour = "steelblue")+
  geom_line(group = 1, colour = "steelblue") +
  labs( x = "\nMonths",
        y = "Number of bike hire",
        title = "Representation of Bike Hire based on the age",
        subtitle = "(Over Months)")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

```

Bike Hire Patterns based on the age & gender

```{r}
 nyc_bikes_df %>% 
  mutate (age = year(start_time) - birth_year) %>% 
  index_by(month) %>% 
  group_by(gender) %>% 
  summarise(age_mean = mean(age, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = age_mean)) +
  geom_col(aes(fill = gender), position = "dodge")+
  labs( x = "\nMonths",
        y = "Age",
        title = "Representation of Bike Hire patterns based on demographics",
        subtitle = "(Data taken from all the months of 2018)")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

```

For each plot you include in your report, please:

**Describe the visualisation: what kind of visualisation is it and what
information does it contain?**

For example: *This scatter plot shows the relationship between age
(years) and spending (GBP), where the red line represents a line of best
fit, and the shaded area represents the 95% confidence interval.*

**Why is this visualisation important to the business? What does it
mean/how can it be used?**

For example: *The highlighted relationship between age and spending can
be used by the business to create better targeted advertisements for
different age groups as well as make better predictions of customer
behaviour.*

# PDA Learning Outcomes

## From Communicating with Data (J4Y7 35)

### 1. Prepare data for visualisation

-   1.1. Domain knowledge
-   1.2. Business requirements
-   1.3. Business processes and concept of data flow
-   1.4. Data visualisation as a tool for decision-making
-   1.5. Data types
-   1.6. Data quality and data bias
-   1.7. Preparing data for visualisation

### 2. Create a range of data visualisations to provide business intelligence

-   2.1. Data visualisation workflow
-   2.2. Data visualisation as a tool for decision-making
-   2.3. Types of data visualisation
-   2.4. Selection of visualisation
-   2.5. Key elements of a data visualisation
-   2.6. Key Performance Indicators (KPIs)
-   2.7. Design of data visualisations
-   2.8. Creating visualisations using software
-   2.9. Misleading visualisations

## From Working with Data (J4Y6 35)

-   4.6 Time series analysis


```{r}
nyc_bikes_df_time <- nyc_bikes_df %>% 
  filter (month %in% c("Jun","Jul","Aug")) %>% 
  mutate (hours =  hour(start_time)) %>% 
  group_by (week, hours) %>% 
  mutate(no_of_bike_hire = n()) 
  
  nyc_bikes_df %>% 
    #filter (month %in% c("Jun","Jul","Aug")) %>%
    filter (month == "Jun") %>% 
   mutate (hours =   hour(start_time)) %>% 
    group_by(week, hours) %>% 
    mutate(no_of_bike_hire = n()) %>% 
  ggplot() + 
  aes(x = week, y = hours ) +
  geom_col(aes(fill = no_of_bike_hire))  +
  geom_line(colour = "steelblue") +
  facet_grid(~month, scales = "free")+
  labs( x = "\nDays",
        y = "Number of bike hire",
        title = "Representation of Bike Hire over months" )+
  scale_x_date(breaks = "1 week", date_labels = "%d")+
  #scale_y_datetime(breaks = "1 hour", date_labels = "%H")+
  color_theme()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

nyc_bikes_df_days
```